---
title: "Welcome!"
---
Hello and welcome to my website! My name is Giulio Orazzo.

I am a finance and data enthusiast with expertise in quantitative modeling, predictive analytics, and data anlysis. I am pursuing a **MS in Quantitative Methods & Modeling** at *CUNY Baruch College- Zicklin School of Business* and hold a **B.S. in Finance** at *CUNY College of Staten Island*. Currently, I work as a Position Control Associate (HR Data Science) at the *Metropolitan Transportation Authority*, where I analyze data, develop predictive models, and create dashboards. Proficient in Python, R, Power BI, VBA, DAX, and STATA, I am passionate about leveraging data-driven insights to solve complex financial and business challenges.
```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```

```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`